D3064R0: Bringing the Laws of Physics, Computer-System Constraints, and Compilers to Bear on the OOTA Problem

### 2024-02-14 ootavacuous.tex
Change "strict C++" to "loose C++"
Since we seem to be drifting in the direction of using "loose C++" for
what Hans Boehm called "strict C++", let's put it in the paper.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

This paper will follow P2055R0~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}
in using the term \emph{full C++} to denote the standard including the
prohibition above, and the term \emph{strict C++}\footnote{
	There is some discomfort with this name, so this footnote
	is reserved for future bikeshedding.}
% @@@ Alternatives thus far: OOTA-prone C++, OOTA-enabled C++,
% @@@ relaxed C++, OOTA C++, OOTA-agnostic C++, loose C++.
to denote a hypothetical standard that excludes this prohibition, but
is otherwise identical to full C++.
Unqualified \emph{C++} means full C++.

@@

Listing~\ref{lst:OOTA Cycle}
shows a code fragment that under strict C++ has an OOTA cycle (although
the cycle is of course prohibited in full C++).
This section describes the fragment, thereby giving an overview of
the \co{herd7} tool.

@@

Under any particular implementation,
a single program can have many different abstract executions,
varying in their decisions about which store each load reads from
and thus the value obtained.
It's worth noting, however, that not all the possible abstract executions
of a program need be realizable by the machine-code executable file
produced by that implementation.
In fact, we will see that \emph{none} of the possible OOTA executions
allowed by the strict C++ abstract machine will ever be realized
by the executables produced by many compilers,

@@

Recall the Simple OOTA example in Listing~\ref{lst:Simple OOTA} on
page~\pageref{lst:Simple OOTA},
in which \co{thread1()} loads the value of \co{x} and stores it in
\co{y} while \co{thread2()} does the reverse.
A globally optimizing compiler given that program might
transform it to the following before translating it into machine code,
if the compiler is sufficiently perverse:
\begin{quote}
\begin{verbatim}
 1 atomic<int> x(0);
 2 atomic<int> y(0);
 3
 4 void thread1()
 5 {
 6   int r1 = 42;
 7   y.store(r1, memory_order_relaxed);
 8 }
 9
10 void thread2()
11 {
12   int r2 = 42;
13   x.store(r2, memory_order_relaxed);
14 }
\end{verbatim}
\end{quote}
The loads previously on lines~6 and~12 have been replaced by constants.
Such a transformation complies with the strict C++ standard,
even though the resulting executable file would produce an unintuitive
OOTA outcome every time it runs!

The only justification a compiler could have for generating output like
this is that it knows exactly what accesses will be performed by both
threads, and therefore it knows that it will not violate the strict C++
memory model by assuming each thread's load reads from the other's
store.
A similar justification can underlie the reasoning in
section~\ref{sec:Semantic Dependencies Affected by Cross-Thread Optimizations};
in principle an analysis of the complete program could lead a compiler
to conclude that \co{y} will always be equal to \co{z} whenever a
particular \co{y - z} expression is evaluated, allowing the compiler
to replace the expression with a constant \co{0}.

@@

We will show that programs built using strict C++ compilers that treat all
atomic operations as volatile can never exhibit OOTA, and the same is true
for compilers obeying the weaker constraint that atomic accesses may not
invented, provided the compiler does not perform multi-thread analysis.

>>>>

This paper will follow P2055R0~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}
in using the term \emph{full C++} to denote the standard including the
prohibition above.
{\it Contra\/} that article, we will use the term \emph{loose C++}\footnote{
	There is some discomfort with this name, so this footnote
	is reserved for future bikeshedding.}
% @@@ Alternatives thus far: OOTA-prone C++, OOTA-enabled C++,
% @@@ relaxed C++, OOTA C++, OOTA-agnostic C++.
(rather than \emph{strict C++}, which seems too similar to
\emph{full C++} for comfort)
to denote a hypothetical standard that excludes this prohibition, but
is otherwise identical to full C++.
Unqualified \emph{C++} means full C++.

@@

Listing~\ref{lst:OOTA Cycle}
shows a code fragment that under loose C++ has an OOTA cycle (although
the cycle is of course prohibited in full C++).
This section describes the fragment, thereby giving an overview of
the \co{herd7} tool.

@@

Under any particular implementation,
a single program can have many different abstract executions,
varying in their decisions about which store each load reads from
and thus the value obtained.
It's worth noting, however, that not all the possible abstract executions
of a program need be realizable by the machine-code executable file
produced by that implementation.
In fact, we will see that \emph{none} of the possible OOTA executions
allowed by the loose C++ abstract machine will ever be realized
by the executables produced by many compilers,

@@

Recall the Simple OOTA example in Listing~\ref{lst:Simple OOTA} on
page~\pageref{lst:Simple OOTA},
in which \co{thread1()} loads the value of \co{x} and stores it in
\co{y} while \co{thread2()} does the reverse.
A globally optimizing compiler given that program might
transform it to the following before translating it into machine code,
if the compiler is sufficiently perverse:
\begin{quote}
\begin{verbatim}
 1 atomic<int> x(0);
 2 atomic<int> y(0);
 3
 4 void thread1()
 5 {
 6   int r1 = 42;
 7   y.store(r1, memory_order_relaxed);
 8 }
 9
10 void thread2()
11 {
12   int r2 = 42;
13   x.store(r2, memory_order_relaxed);
14 }
\end{verbatim}
\end{quote}
The loads previously on lines~6 and~12 have been replaced by constants.
Such a transformation complies with the loose C++ standard,
even though the resulting executable file would produce an unintuitive
OOTA outcome every time it runs!

The only justification a compiler could have for generating output like
this is that it knows exactly what accesses will be performed by both
threads, and therefore it knows that it will not violate the loose C++
memory model by assuming each thread's load reads from the other's
store.
A similar justification can underlie the reasoning in
section~\ref{sec:Semantic Dependencies Affected by Cross-Thread Optimizations};
in principle an analysis of the complete program could lead a compiler
to conclude that \co{y} will always be equal to \co{z} whenever a
particular \co{y - z} expression is evaluated, allowing the compiler
to replace the expression with a constant \co{0}.

@@

We will show that programs built using loose C++ compilers that treat all
atomic operations as volatile can never exhibit OOTA, and the same is true
for compilers obeying the weaker constraint that atomic accesses may not
invented, provided the compiler does not perform multi-thread analysis.
